\texttt{MSnbase}, efficient R-based access and manipulation of raw mass spectrometry data

### 2020-04-27 msnbase2.Rnw
add discussion

In this software note, we present version 2 of the \texttt{MSnbase}
\cite{Gatto:2012} R/Bioconductor software package. \texttt{MSnbase}
offers a platform that lies between low level libraries and end-use
software. It provides a flexible command line environment for
metabolomics and proteomics mass spectrometry-based application, that
allows a detailed step-by-step processing, analysis and exploration of
the data and development of novel computational mass spectrometry
methods.

@@

The R environment is well known for the quality of its visualiation
capacity. This also holds true for mass
spectrometry\cite{Gatto:2015,protViz,Pviz,Bemis:2015}. Here, we
conclude the overview of version 2 of the \texttt{MSnbase} package by
highlighting the flexibility of the software to visualise and assess
the efficiency of raw data processing. Figure \ref{fig:cent} compares
the raw MS profile data for the serine and the same data after
smoothing, centroiding and M/Z refinement.

@@

To address (from guidelines):

\begin{itemize}

\item{potential for reuse}: see
  \cite{Smith:2006,Wieczorek:2017,Griss:2019} for examples.
\item{general limitations}
\item{system limitations}
\item{end-user documentation}
\item{developer documentation}
\item{sample data}
\item{benchmark data set}
\item{availability}
\item{license information}
\item{system requirements}

\end{itemize}

Collaborative development, 11 contributors since creation (see blog
post).

Count packages depending on \texttt{MSnbase}.

@@

The version of \texttt{MSnbase} used in this manuscritp is version
2.10.0. The main features presented here
were available since version 2.0.

>>>>

In this software note, we present version 2 of the \texttt{MSnbase}
\cite{Gatto:2012} software package, available from the
Bioconductor~\cite{Huber:2015} project since 2011. \texttt{MSnbase}
offers a platform that lies between low level libraries and end-use
software. It provides a flexible command line environment for
metabolomics and proteomics mass spectrometry-based application, that
allows a detailed step-by-step processing, analysis and exploration of
the data and development of novel computational mass spectrometry
methods.

@@

The R environment is well known for the quality of its visualiation
capacity. This also holds true for mass
spectrometry\cite{Gatto:2015,protViz,Pviz,Bemis:2015}. Here, we
conclude the overview of version 2 of the \texttt{MSnbase} package by
highlighting the flexibility of the software to visualise and assess
the efficiency of raw data processing. Figure \ref{fig:cent} compares
the raw MS profile data for the serine and the same data after
smoothing, centroiding and M/Z refinement. Detailed execution and
description of these operations can be found in the \emph{MSnbase:
  centroiding of profile-mode MS data} \texttt{MSnbase} vignette.

@@

The first public commit to the \texttt{MSnbase} GitHub repository was
in October 2010. Since then, the package benefited from 12
contributors\cite{contribs} that added various features, some
particularly significant ones such as the on-disk backend described
above. According to the package's Bioconductor page, there are 36
Bioconductor packages that depend, import or suggest it. Among these
are \texttt{pRoloc}~\cite{Gatto:2014a}, \texttt{DAPAR} and
\texttt{ProStaR}~\cite{Wieczorek:2017},
\texttt{RMassBank}~\cite{Stravs:2013},
\texttt{msmsTests}~\cite{msmsTests}, \texttt{DEP}~\cite{Zhang:2018},
\texttt{MSstatsQC}~\cite{Dogu:2017} and the widerly used
\texttt{xcms}~\cite{Smith:2006}. \texttt{MSnbase} is also used in
packages that are not on Bioconductor, such as
IsoProt~\cite{Griss:2019}, for examples. The package currently ranks
101 out of 1823 packages based on the monthly downloads from unique IP
addresses, tallying over 1000 downloads from unique IP addresses each
months.

As is custom with Bioconductor packages, \texttt{MSnbase} comes with
ample documentation. Every user-accessible function is documented in a
dedicated manual page. In addition, the package offers 5 vignettes,
including one aimed at developers. Question from users and developers
are answered on the Bioconductor support forum as well as on the
package GitHub page. The package provides several sample and
benchmarking datasets, and relies on other dedicated
\textit{experiment packages} such as \texttt{msdata}~\cite{msdata} for
raw data or \texttt{pRolocdata}~\cite{Gatto:2014a} for quantiative
data.

\texttt{MSnbase} is available on Windows, Mac OS and Linux under the
open source Artistic 2.0 license and easily installable using standard
installation procedures.

@@

<<version>>=
v <- packageVersion("MSnbase")
@

The version of \texttt{MSnbase} used in this manuscritp is version
\Sexpr{v}. The main features presented here were available since
version 2.0.
